% Open Problems — External Resources
% Draftable now from research cards' open questions sections

\section{Open Problems}
\label{sec:open-questions}

The following problems remain unresolved across the SNEA sources. These are
drawn from the ``Open Questions for Linguist Review'' sections of each
research card.

\subsection{Phonological}

\begin{enumerate}
  \item \textbf{Preaspiration inference rules:} Can a reliable set of
    morphotactic rules predict when preaspiration /h/ was present but
    unwritten? Or does this require per-entry comparative verification?
  \item \textbf{The \textit{gh} grapheme:} Is Edwards' \textit{gh} /x/, /h/,
    or /k/? Cross-linguistic cognate evidence from Munsee Delaware is needed.
  \item \textbf{Nasalization disambiguation:} When does \textit{an}/\textit{aun}
    represent /ã/ vs.\ /an/? Massachusett cognate comparison is the primary
    method, but it is not always available.
\end{enumerate}

\subsection{Methodological}

\begin{enumerate}
  \item \textbf{Pidgin vs.\ genuine dialect:} Are the simplifications in
    Winslow's data evidence of a pre-existing Massachusett Pidgin (Goddard
    1977) or simply a learner's imperfect recording?
  \item \textbf{Dialect stratification:} How much of the variation between
    Natick (Eliot) and Pokanoket (Winslow/Wood) is orthographic vs.\
    phonological vs.\ dialectal?
  \item \textbf{Fielding circularity:} Which Fielding (2013) entries derive
    from Prince-Speck vs.\ independent sources? A source-tracing audit is
    needed.
\end{enumerate}

\subsection{Corpus-Level}

\begin{enumerate}
  \item \textbf{Training data shortage:} Only 3 of 2,134 Narragansett records
    have phonetic fields. Hand-annotating \textasciitilde50 entries is
    recommended before ML approaches are viable.
  \item \textbf{Tiny corpus reliability:} Winslow's 24 records --- can a
    dataset this small yield reliable phonological data?
\end{enumerate}
